Eco-evolutionary dynamics of finite populations from first principles

### 2023-03-17 MS_thesis.tex
Make bibliography 2 column
Revert to single column bibliography because 2 column looks terrible in retrospect
Update abstract

\documentclass[twoside, 12pt,lot,lof]{iiser-thesis-modified} %Use the custom .cls file we have

@@

%The abstract goes here. IISER guidelines say it should be <= 250 words.
\thesisabstract{
Population biology is built on a strong mathematical foundation developed during the Modern Synthesis and encapsulated in fields such as population genetics, evolutionary game theory, and quantitative genetics. Historically, many of these formalisms have often worked with infinite populations, ignoring the effects of demographic stochasticity. Finite population models in population and quantitative genetics usually assume a fixed population size and are of limited applicability in the real world, where population sizes routinely fluctuate. In this thesis, I use ideas from statistical physics to analytically describe evolution in finite, fluctuating populations from biological first principles. Starting from a density-dependent `birth-death process’ describing an arbitrary closed population of individuals with discrete traits, I derive a set of stochastic differential equations (SDEs) for how trait frequencies change over time. Along with recovering the effects of the standard evolutionary forces of selection, mutation, and drift, these SDEs also reveal a new directional evolutionary force, `noise-induced selection’, that is particular to finite, fluctuating populations and has therefore been largely overlooked in standard formulations of evolution. Noise-induced selection can cause deviations from neutrality and can reverse the direction of evolution predicted by infinite-population frameworks, with implications for simulation studies and real-world populations. Well-known results such as the replicator-mutator equation and Fisher’s fundamental theorem are recovered in the infinite population limit. Finally, I extend these ideas to one-dimensional quantitative traits through a `stochastic field theory’ that yields frameworks such as Kimura's continuum-of-alleles and Lande's gradient dynamics in the infinite population limit. My work thus generalizes the formal structures of population biology to finite, fluctuating populations and predicts a new evolutionary force unique to such populations.
}

@@

\pagestyle{plain} %Don't make the initial frontmatter fancy
\thesisfront
%\listoftables
%\addcontentsline{toc}{chapter}{List of Tables}
%\listoffigures
%\addcontentsline{toc}{chapter}{List of Figures}

@@

%You need the below commands to add a 'Part' level reference to the Appendices in the TOC
%from https://tex.stackexchange.com/a/286452
\cleardoublepage
\phantomsection
\appendix
\addcontentsline{toc}{part}{Appendices}

@@

%Import the Appendices
\chapter{From It\^{o} to Fokker-Planck}\label{App_SDE_FPE}
\input{ind_files/99_Appendix_A}
\chapter{Deriving stochastic trait frequency dynamics using It\^{o}'s formula}\label{App_density_to_freq}
\input{ind_files/99_Appendix_B}
\chapter{A Price-like equation for the variance of a type-level quantity}\label{App_stoch_var_eqns}
\input{ind_files/99_Appendix_C}
\chapter{Some Examples}\label{App_examples}
\input{ind_files/99_Appendix_D_examples}

%Add the bibliography and make an entry in the table of contents
\cleardoublepage
\phantomsection
\printbibliography
\addcontentsline{toc}{part}{References}

>>>>

\documentclass[twoside,12pt]{iiser-thesis-modified} %Use the custom .cls file we have with two sided paper and 12pt font

@@

%The abstract goes here. IISER guidelines say it should be <= 250 words.
\thesisabstract{
Population biology is built on a strong mathematical foundation developed during the Modern Synthesis and encapsulated in fields such as population genetics, evolutionary game theory, and quantitative genetics. Historically, many of these formalisms have often worked with infinite populations, ignoring the effects of demographic stochasticity. Finite population models in population and quantitative genetics usually assume a fixed population size and are of limited applicability in the real world, where population sizes routinely fluctuate. In this thesis, I use ideas from statistical physics to analytically describe evolution in finite, fluctuating populations from biological first principles. Starting from a density-dependent `birth-death process’ describing an arbitrary closed population of individuals with discrete traits, I derive a set of stochastic differential equations (SDEs) for how trait frequencies change over time. These SDEs reveal a new directional evolutionary force, `noise-induced selection’, that is particular to finite, fluctuating populations and has therefore been largely overlooked in standard formulations of evolution. Noise-induced selection can cause deviations from neutrality and can reverse the direction of evolution predicted by infinite-population frameworks. These SDEs also recover well-known results such as the replicator-mutator equation and Fisher’s fundamental theorem in the infinite population limit, and can help explain and organize several previous studies under a single theoretical framework. They also have implications for empirical studies and the formulation of evolutionary agent-based models. Finally, I extend the birth-death formalism to one-dimensional quantitative traits through a `stochastic field theory’ that yields frameworks such as Kimura's continuum-of-alleles and Lande's gradient dynamics in the infinite population limit. My work thus generalizes the formal structures of population biology to finite, fluctuating populations and predicts a new evolutionary force unique to such populations.
}

@@

\pagestyle{plain} %Don't make the initial frontmatter fancy

%Add the frontmatter
\thesisfront

@@

%You need the below commands to add a 'Part' level reference to the Appendices in the TOC
%from https://tex.stackexchange.com/a/286452
%The \phantomsection is needed for hyperref to jump to the correct location
\cleardoublepage
\phantomsection
\appendix
\addcontentsline{toc}{part}{Appendices}

@@

%Import the Appendices
\chapter{From It\^{o} to Fokker-Planck}\label{App_SDE_FPE}
\input{ind_files/99_Appendix_A}
\chapter{Deriving stochastic trait frequency dynamics using It\^{o}'s formula}\label{App_density_to_freq}
\input{ind_files/99_Appendix_B}
\chapter{A Price-like equation for the variance of a type-level quantity}\label{App_stoch_var_eqns}
\input{ind_files/99_Appendix_C}
\chapter{Some Examples}\label{App_examples}
\input{ind_files/99_Appendix_D_examples}ne

%Bibliography
%Again, add the phantomsection so that hyperref knows where to go
\cleardoublepage
\phantomsection
\Huge\printbibheading
%Putting \printbibheading up here makes sure the heading is printed before I change the font size below.
%Change the font for the bibliography
\renewcommand*{\bibfont}{\footnotesize}	
%Add the bibliography
\printbibliography[heading=none]
%Add an entry in the table of contents
\addcontentsline{toc}{part}{References}
